Natural sources of uncertainty and their implications

### 2017-04-28 book_chapter/paper.tex
reference game section edits

\section{Vagueness in referential games}
\label{sec:referential-vagueness}

Linguistic interaction, as captured by a sim-max game, has a rather narrow scope. There is a fixed prior and fixed utility function and a fixed similarity measure between states. If the sender observes a state $t$ and the receiver guesses state $t'$ the utility of that interaction is proportional to the similarity between $t$ and $t$. Here is a natural example of  a situation captured by a sim-max model: Kiki wants Bubu to draw her dissertation cover in a particular shade of blue $t$; the closer Bubu's color choice matches Kiki's idea, the better. We could say that Kiki wants to describe the color for Bubu to pick out. It's a natural thing to want to do with language. But a lot of communication with vague and contiguously variable properties is rather different. In the general spirit of Wittgenstein's language games, we should look at some more cases of language use side by side. Let us therefore consider a continuum of cases of referential communication.

@@

Interestingly, referential communication can take place under a plethora of different epistemic conditions. A lot of modeling work considers \emph{reference games without displacement} where sender and receiver communicate about a context, i.e. a set of objects, which is immediately accessible in their direct perceptual environment \parencite{BaronchelliGong2010:Modeling-the-Em,Franke2012:Scales-Salience,Franke2012:On-Scales-Salie}.\footnote{add more references later!} Agents may make mistakes about observing exactly how blue or long etc.~a particular object is, but there is no further fundamental uncertainty about the context. This, of course, need not be so. \textcite{lipman_why_2009} and \textcite{Deemter2009:Utility-and-Lan}, for example, consider \emph{reference games with displacement}. Kiki asks Bubu to fetch the red book from the library. No other book than the one Kiki needs will serve. Yet when Kiki describes the book, the perceptual context in which Bubu must make her choice is not co-present to Kiki. Yet other situation are imaginable. In \emph{reference games with partial uncertainty}, for instance, interlocutors may know that they have shared and immediate access only to a subset of the objects in the context, but be uncertain about whether the co-player sees any other objects. This is the set-up for the classic director task in psycholinguistics \parencite[\emph{e.g.}][]{KraussGlucksberg1977:Social-and-nons,Keysar2000,KeyzarLin2003:Limits-on-Theor}.

>>>>

\section{Vagueness in multiple reference games}
\label{sec:referential-vagueness}

Linguistic interaction, as captured by a sim-max game, has a rather narrow scope. There is a fixed prior, a fixed utility function and a fixed similarity measure between states. If the sender observes a state $t$ and the receiver guesses state $t'$ the utility of that interaction is proportional to the similarity between $t$ and $t$. Here is a natural example of  a situation captured by a sim-max model: Kiki wants Bubu to draw her dissertation cover in a particular shade of blue $t$; the closer Bubu's color choice matches Kiki's idea, the better. We could say that Kiki wants to \emph{describe} the color for Bubu to pick out. It's a natural thing to want to do with language. But a lot of communication with vague and contiguously variable properties is rather different. In the general spirit of Wittgenstein's method of arranging multiple language games side by side for a more encompassing picture, we should look at some more cases of language use as well. Let us therefore consider a continuum of cases of referential communication.

@@

Interestingly, referential communication can take place under a plethora of different epistemic conditions. A lot of modeling work considers \emph{reference games without displacement} where sender and receiver communicate about a context, i.e. a set of objects, which is immediately accessible in their direct perceptual environment \parencite{BaronchelliGong2010:Modeling-the-Em,Franke2012:Scales-Salience,Franke2012:On-Scales-Salie}.\footnote{add more references later!} Agents may make mistakes about observing exactly how blue or long etc.~a particular object is, but there is no further fundamental uncertainty about the context. This, of course, need not be so. \textcite{lipman_why_2009} and \textcite{Deemter2009:Utility-and-Lan}, for example, consider \emph{reference games with displacement}. Kiki asks Bubu to fetch the red book from the library. No other book than the one Kiki needs will serve. Yet when Kiki describes the book, the perceptual context in which Bubu must make her choice is not co-present to Kiki. Yet other situations are imaginable. In \emph{reference games with partial uncertainty}, for instance, interlocutors may know that they have shared and immediate access only to a subset of the objects in the context, but be uncertain about whether the co-player sees any other objects. This is the set-up for the classic director task in psycholinguistics \parencite[\emph{e.g.}][]{KraussGlucksberg1977:Social-and-nons,Keysar2000,KeyzarLin2003:Limits-on-Theor}.

What would a rational agent do in repeated reference games? Take a universe in which agents play repeated reference games without displacement. A single game has sender and receiver both flawlessly perceive a context of $k$ objects (with $k$ possibly randomly drawn for each encounter); each object has its individual features, e.g., drawn from a rich multidimensional feature space of continuously variable properties \parencite[\emph{e.g.}][]{Franke2012:Scales-Salience,Franke2012:On-Scales-Salie}. The sender chooses a message $m$ for the target object $t$, but she can condition her choice on the current context $c$. Likewise, the receiver chooses a referent $t'$ from the available objects in $c$, again conditional on $c$. But if agents play many such reference games with different contexts $c$ and $c'$, and if strategic rationality means to signal and interpret so as to coordinate on the correct referent, strategic rationality of local choices in $c$ and $c'$ is not enough to guarantee that the agents' beliefs and behavior in $c$ have any relation to their beliefs and behavior in $c'$. In other words, local strategic rationality in a fixed context does not entail a globally consistent behavior across multiple contexts. The situation only gets worse if we want to have agents use the same expressions in a consistent way also across different types of reference games, with possibly different epistemic conditions.

The picture sketched in the last paragraph neglects a rational agent's ability to learn from previous experience and to generalize from prior experience. Which strategies for generalization are best depends on the stochastic properties of the environment in which learning and decision making takes place. The problem for language is that it what works best for on agent depends on what the other agent does. Communication problems are games in the formal sense of delivering payoffs only as a function of what all players do. Consequently, the problem of how best to adapt behavior from past experience in language games also depends on how other agents categorize, generalize and behave. Nothing gets off the ground with hyper-rationality alone. There needs to be learning, evolution and circumstances. As with sim-max games above, biases and other quirks of general cognition are likely only helpful in solving this massive coordination problem.

\begin{itemize}
\item uncertainty
\end{itemize}

What about meaning and vagueness in a universe of multiple language games and heavy uncertainty? We could look for meaning at several levels. Firstly, as before, there is the actual behavior of a single agent in each actualized language game. Like with sim-max games, behavior that strictly maximizes expected utility under uncertainty may be resource heavy, so it might be compatible with rationality that agents' production choices are stochastic (with the caveat that in the real world we would never know because we cannot ever elicit a production in the exact same circumstances in a constantly changing environment including the mind of a speaker). Secondly, if we look at behavior across many game types and contexts, there is also the level of an agent's internal theory of how words and phrases are likely to be used, conditional on a given context. Either one of these two levels, the agent's behavior or linguistic generalizations, could be the foundation for a notion of emerging abstract meaning as constituted by the practices of a population at a particular time. The question is: for which of these three options of situating meaning would we like to have a theory of so that being vague would be disastrous for that theory?
